---
title: "Introduction to R and RStudio"
toc: true
toc-location: right
---

```{r}
#| echo: false
library(tidyverse)
```

# Introduction

-   Finding your way round RStudio
-   Typing in data and plotting it
-   Importing data: working directories and paths

# Finding your way round RStudio

## RStudio: live demonstration

-   the panels
    -   Top left: Script - write and edit code and comments to keep
    -   Bottom left: Console - where commands get executed and can be
        typed
    -   Top right: Environment - where you can see the objects you have
        created; History - of commands
    -   Bottom right: Files - a file explorer; Plots; Packages; Help\
-   making yourself comfortable
-   typing in the console sending commands
-   using R as a calculator
-   assigning values
-   where to see objects
-   using a script - make sure to execute
-   comments \#
-   data types and structures
-   functions `c()`
-   types of R files: `.R`, `.RData` `.RHistory`

## Overview of demonstration

There are several ways you can *recap* the demo at a later date:

🖼️ Refer to this infographic [Full size RStudio
Anatomy](https://raw.githubusercontent.com/UKRI-Digital-Research-Skills-Catalyst/introduction-to-reproducibility-with-r/80b47b8f3ea698ee3855afe077a8e12f2569070b/materials/images/RStudio%20Anatomy.svg)

![infographic](images/RStudio%20Anatomy.svg){width="600"}

📖 Read [First Steps in
RStudio](https://3mmarand.github.io/comp4biosci/first_steps_rstudio.html)

📹 Watch

{{< video https://www.youtube.com/watch?v=7oLLjkTNrjY >}}

{{< video https://youtu.be/2Ke1FGdaMSI >}}

# Typing in data and plotting it

🎬 We code together

🛝 You try!

## The goal 🐈 plot

We will work with some data on the coat colour of 62 cats. You are going
to type data in R, summarise and plot it

The data are as a frequency table:

```{r}
#| echo: false
coat <- c("black", 
          "white", 
          "tabby", 
          "ginger", 
          "tortoiseshell", 
          "calico")
freq <- c(23, 15, 8, 10, 5, 1)
df <- data.frame(coat, freq)
knitr::kable(df, 
             table.attr = "style='width:30%;'",
             caption = "Frequency of coat colours in 62 cats",
             format = "html",
             col.names = c("Coat colour", "No. cats"),
             row.names = FALSE) |>  
  kableExtra::kable_styling()
```

You will create a figure like this:

```{r}
#| echo: false
coat_data <- data.frame(coat, freq)
ggplot(data = coat_data, 
       aes(x = coat, y = freq)) +
  geom_col(colour = "black", fill = "lightblue") +
  scale_x_discrete(name = "Coat colour",
                   expand = c(0, 0)) +
  scale_y_continuous(name = "Number of cats",
                     expand = c(0, 0),
                     limits = c(0, 25)) +
  theme_classic()
```

## Getting set up

🎬 In RStudio do File \| New project \| New directory. Be purposeful
about where you create it and name it. I suggest `r-july`

🎬 Make a new script and save it as `type-data-and-plot.R` to carry out
the rest of the work.

## Creating the data

Start by making a vector called `coat` that holds coat colours

🎬 Write the following in your script:

```{r}
# coat colours
coat <- c("black", "white", "tabby", "ginger", "tortoiseshell", "calico")
```

The shortcut for `<-` is Alt+- (hold the Alt key down then hit the minus
key ).

Ensure your cursor is on the line with the command and do Control+Enter
to send the command to the console to be executed.

I have used a comment. Comment your code as much as possible!

🛝 Create a vector called `freq` containing the numbers of cats with
each coat colour

```{r}
#| code-fold: true
#| code-summary: "Answer - don't look until you have tried!"

# numbers of cats with each coat colour
freq <- c(23, 15, 8, 10, 5, 1)

```

## Packages

Commands like `c()` and `sum()` are in packages which are part the
'base' R system. A package is a collection of related commands. Base
packages are installed automatically when you install R.

Other packages, such as **`ggplot2`** [@ggplot2] need to be *installed
once*, and then *loaded each session*. **`ggplot2`** is one of the
**`tidyverse`** [@tidyverse] packages.

🎬 Load the **`tidyverse`**:

```{r}
#| eval: false
library(tidyverse)
```

You will likely be warned of some function name conflicts but these will
not be a problem for you.

## Plotting the data with `ggplot()`

-   `ggplot()` takes a dataframe for an argument
-   We can make a dataframe of the two vectors, `coat` and `freq` using
    the `data.frame()` function.

🎬 Make a dataframe called `coat_data`

```{r}
coat_data <- data.frame(coat, freq)
```

🎬 Click on `coat_data` in the Environment to open a spreadsheet-like
view of it.

🎬 Create a simple barplot using `ggplot` like this:

```{r}
ggplot(data = coat_data, aes(x = coat, y = freq)) +
  geom_col()
```

-   `ggplot()` alone creates a blank plot.
-   `ggplot(data = coat_data)` looks the same.
-   `aes()` gives the 'Aesthetic mappings'. How variables (columns) are
    mapped to visual properties (aesthetics) e.g., axes, colour, shapes.
-   Thus `ggplot(data = coat_data, aes(x = coat, y = freq))` produces a
    plot with axes
-   `geom_col`: A 'Geom' (Geometric object) gives the visual
    representations of the data: points, lines, bars, boxplots etc.

Note that `ggplot2` is the name of the package and `ggplot()` is its
most important command.

## Using the help manual

The manual tells us about a function. 'Arguments' can be added to the
`geom_col()` command inside the brackets.

Commands do something and their arguments (in brackets) and can specify:

-   what object to do it to
-   how exactly to do it

Many arguments have defaults so you don't always need to supply them.

🎬 Open the manual page for `geom_col()` using:

```{r}
#| eval: false
?geom_col
```

The manual page has several sections.

-   **Description** an overview of what the command does
-   **Usage** lists arguments
    -   form: argument name = default value
    -   some arguments MUST be supplied others have defaults
    -   `...` means etc and includes arguments that can be passed to
        many 'geoms'
-   **Arguments** gives the detail about the arguments
-   **Details** describes how the command works in more detail
-   **Value** gives the output of the command
-   Don't be too perturbed by not fully understanding the information

> **Aesthetics** geom_col() understands the following aesthetics.
> Required aesthetics are displayed in bold and defaults are displayed
> for optional aesthetics:
>
> -   x
> -   y
> -   alpha → NA
> -   colour → via theme()
> -   fill → via theme()
> -   group → inferred
> -   linetype → via theme()
> -   linewidth → via theme()
> -   width → 0.9

## Customising a plot

### Bar colour

🎬 Change the fill of the bars using `fill`:

```{r}

ggplot(data = coat_data, aes(x = coat, y = freq)) +
  geom_col(fill = "lightblue")
```

Colours can be given by their name, "lightblue" or code, "#ADD8E6".

Look up by [name](assets/colournames.pdf) or
[code](assets/colourhex.pdf)

🎬 Change the bars to a colour you like.

`fill` is one of the arguments covered by `...`. `fill` is an
'aesthetic'. If you look for `...` in the list of arguments you will see
it says:

> Other arguments passed on to layer(). These are often aesthetics, used
> to set an aesthetic to a fixed value, like colour = "red" or size = 3.
> They may also be parameters to the paired geom/stat.

We just set the `fill` aesthetic to a fixed value.

Further down the manual, there is a section on **Aesthetics** which
lists those understood by `geom_col()`

We can set (map) the `fill` aesthetic to a fixed colour inside
`geom_col()` *or* map it to a variable from the dataframe inside the
`aes()` instead. This means the colour will be different for different
values in that variable.

🎬 Map the `fill` aesthetic to the `coat` variable:

```{r}
ggplot(data = coat_data, aes(x = coat, y = freq, fill = coat)) +
  geom_col()
```

Note that we have taken `fill = "lightblue"` out of the `geom_col()` and
instead put `fill = coat` in the `aes()`.

🛝 Use the manual to change the colour of the lines around each bar to
black.

```{r}
#| code-fold: true
#| code-summary: "Answer - don't look until you have tried!"

ggplot(data = coat_data,
       aes(x = coat, y = freq)) +
  geom_col(fill = "lightblue", 
           colour = "black")
```

### Changing the axes

We can make changes to the axes using:

-   Changes to a discrete x axis: `scale_x_discrete()`
-   Changes to a continuous y axis: `scale_y_continuous()`

`ggplot` automatically extends the axes slightly. You can turn this
behaviour off with the `expand` argument in `scale_x_discrete()` and
`scale_y_continuous()`.[^1]

[^1]: There are also `scale_x_continous()` and `scale_y_discrete()`
    functions when you have those types of variable

🎬 To remove the gap between the axes and the data:

```{r}
ggplot(data = coat_data, aes(x = coat, y = freq)) +
  geom_col(fill = "lightblue", 
           colour = "black") +
  scale_x_discrete(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0)) 
```

Each 'layer' is added to the ggplot() command with a `+`

::: callout-tip
## Top Tip

Make your code easier to read by using white space and new lines

-   put spaces around `=` , `->` and after `,`
-   use a newline after every comma in a command with lots of arguments
:::

🛝 Look up `scale_x_discrete` in the manual and work out how to change
the axis title from "coat" to "Coat colour". Also change the *y*-axis
title.

```{r}
#| code-fold: true
#| code-summary: "Answer - don't look until you have tried!"

ggplot(data = coat_data, aes(x = coat, y = freq)) +
  geom_col(fill = "lightblue", 
           colour = "black") +
  scale_x_discrete(expand = c(0, 0),
                   name = "Coat colour") + 
  scale_y_continuous(expand = c(0, 0),
                     name = "Number of cats") 


```

🎬 I would prefer to see the *y*-axis extend a little beyond the data
and we can change the axis "limits" in the `scale_y_continuous()`

```{r}
ggplot(data = coat_data, aes(x = coat, y = freq)) +
  geom_col(fill = "lightblue", 
           colour = "black") +
  scale_x_discrete(expand = c(0, 0),
                   name = "Coat colour") + 
  scale_y_continuous(expand = c(0, 0),
                     name = "Number of cats",
                     limits = c(0, 25)) 


```

### Getting rid of the grey background

The grey grid background is useful for examining plots on a screen but
for a report or publication you will want a more scientific style. Every
aspect of the "theme" of a plot - the non-data elements such as fonts,
background colours, axis line colours etc - can be controlled
individually[^2] but there are some handy built in themes that apply
several changes at once. One of these is `theme_classic()`

[^2]: [Modify components of a
    theme](https://ggplot2.tidyverse.org/reference/theme.html)

🎬 Add `theme_classic()` to the plot:

```{r}
ggplot(data = coat_data, aes(x = coat, y = freq)) +
  geom_col(colour = "black",
           fill = "lightblue") +
  scale_x_discrete(expand = c(0, 0),
                   name = "Coat colour") + 
  scale_y_continuous(expand = c(0, 0),
                     name = "Number of cats",
                     limits = c(0, 25)) +
  theme_classic()


```

# Summary

## 🔧 RStudio Interface

-   Script – Where you write and save your code (.R or .qmd files)
-   Console – Where commands are executed and immediate output appears
-   Environment – Lists your objects and data
-   Files / Plots / Help – Where you can manage files, view plots, read
    help docs

## 🧠 Core ideas

-   Assignment – Use `<-` to store values: `x <- 5`
-   Vectors – One-dimensional data: `c(1, 2, 3)`
-   Dataframes – Table-like structures that are most widely used:
    `data.frame(name = c("A", "B"))`
-   Functions – Commands that perform actions: `mean(x)`, `ggplot()`
-   Data import functions read data from files to dataframes for
    analysis
-   Packages like **`tidyverse`** are loaded with `library()`
-   Always use RStudio Projects and paths relative to the project folder
    so your code is reproducible and portable

## 📊 Plotting with **`ggplot2`**

-   Structure – `ggplot(data, aes(x, y)) + geom_*()`
-   Common geoms – `geom_col()`, `geom_boxplot()`, `geom_point()`
-   Customize – `fill`, `color`, `width`, `theme_classic()`
-   Scales – Adjust axes with `scale_*_continuous()` or
    `scale_*_discrete()`

Pages made with R [@R-core], Quarto [@Allaire_Quarto_2026], `knitr`
[@knitr1; @knitr2; @knitr3], `kableExtra` [@kableExtra]

# References
